---
title: "Final_GWAS data analysis and meta -analysis"
author: "samuel"
date: "2026-05-15"
output: html_document
---

```{r}
#load tools

library(ggplot2)
library(qqman)
library(dplyr)
library(ggrepel)


```


```{r}
#load GWAS output data
Log23_100 <- read_table("~/Gwas_final/final_pc23_gwas.assoc.logistic")


```

```{r}
#plotting a qqplot 

qq(Log23_100$P)
```


```{r}
# preparing the data for manhattan plot

Logpc23_100 <- Log23_100 %>% 
  
  # Compute chromosome size
  group_by(CHR) %>% 
  summarise(chr_len=max(BP)) %>% 
  
  # Calculate cumulative position of each chromosome
  mutate(tot=cumsum(chr_len)-chr_len) %>%
  select(-chr_len) %>%
  
  # Add this info to the initial dataset
  left_join(Log23_100, ., by=c("CHR"="CHR")) %>%
  
  # Add a cumulative position of each SNP
  arrange(CHR, BP) %>%
  mutate( BPcum=BP+tot) %>%

  # Add highlight and annotation information
  #mutate(is_highlight = ifelse(-log10(P) > 7.3, "yes", "no")) %>%
  mutate( is_annotate=ifelse(-log10(P)>7.3, "yes", "no")) 

# Prepare X axis
axisdf <- Logpc23_100 %>% group_by(CHR) %>% summarize(center=( max(BPcum) + min(BPcum) ) / 2 )
```


```{r}
#ploting ggplot manhattan
ggplot(Logpc23_100, aes(x = BPcum, y = -log10(P))) +
  geom_point(aes(color = as.factor(CHR)), alpha = 0.8, size = 1.3) +
  scale_color_manual(values = rep(c("brown", "skyblue"), 22)) +
  scale_x_continuous(labels = axisdf$CHR, breaks = axisdf$center) +
  scale_y_continuous(expand = c(0, 0),limits = c(0, 12.5)) +
  geom_hline(yintercept = 7.3, color = "red", linetype = "dashed") +  # e
  #geom_hline(yintercept = 8, color = "blue", linetype = "dashed") + #lowerlimit
  theme_bw() +
   # Add highlighted points
  #geom_point(data=subset(metaan2, is_highlight=="yes"), color="orange", size=2) +
  
    # Add label using ggrepel to avoid overlapping
  geom_label_repel( data=subset(Logpc23_100, is_annotate=="yes"), aes(label=SNP), size=2) +
  
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  xlab("Genomic Position") +
  ylab("-log10(p-value)")

```


```{r}
#bonferoni correction
n_tests <- nrow(done)  # or replace with the total number of SNPs if different

# Desired significance level
alpha <- 0.05

# Bonferroni correction is standard of significance
bonferroni_threshold <- alpha / n_tests
bonferroni_threshold
log10_threshold <- -log10(bonferroni_threshold)
log10_threshold #upper limit


#lower limit
alpha <- 0.05
lowerthreshold = alpha/1000000
threshold = -log10(lowerthreshold)


```



```{r}
#calculating additional parameters
Logpc23_100$beta<- log(Logpc23_100$OR)
Logpc23_100$SE <- Logpc23_100$beta/Logpc23_100$STAT

```

```{r}
#including sample size
Logpc23_100$N <- 107
```



```{r}
#Preparing GWAMA tables
gwama_cev <- filtered_gwas2 %>%
  mutate(
    # Marker name
    MARKERNAME = SNP,
    
    # Effect allele
    EA = A1,
    
    # Other allele
    NEA = A2,
    
    # Effect allele frequency (make sure this exists in your data)
    EAF = F_U,
    
    # Effect (OR) – if you only have beta, convert if needed
    OR = OR,
    
    # 95% CI
    OR_95U = exp(beta + 1.96 * SE),
    OR_95L = exp(beta - 1.96 * SE),
    
    # Strand
    STRAND = "+",
    
    # Sample size
    N = N,
    
    # Chromosome
    CHROMOSOME = CHR,
    
    # Position
    POSITION = BP
  ) %>%
  select(
    MARKERNAME,
    EA,
    NEA,
    EAF,
    OR,
    OR_95U,
    OR_95L,
    STRAND,
    N,
    CHROMOSOME,
    POSITION
  )

write.csv(gwama_clean, "E:/Gwas_final/cev_input.csv", row.names = FALSE)
```




```{r}
#gwama_clean <- na.omit(gwama_cev)

# Write out to a CSV for GWAMA (if needed)

```



```{r}
#loading GWAS summary statistics from GWAS catalog
metastat_1 <- read_table("E:/Gwas/GWAMA/GCST012776.h.tsv.gz")
metastat_2 <- read_table("E:/Gwas/GWAMA/34594039-GCST90018817-EFO_0001061.h.tsv.gz")
metasta_3 <- read_table("E:/Gwas/GWAMA/34594039-GCST90018597-EFO_0001061.h.tsv.gz")
metasta_4 <- read_table("E:/Gwas/GWAMA/34737426-GCST90043929-MONDO_0042487.h.tsv.gz")


```


```{r}
#Preparing gwamma input for summary statistics

metastat_1$N <- 1224

#prepare gwas_cat meta 1
meta1 <- metastat_1 %>%
  mutate(
    # Marker name
    MARKERNAME = rsid,   # or SNP_id if preferred
    
    # Effect allele
    EA = effect_allele,
    
    # Other allele
    NEA = other_allele,
    
    # Effect allele frequency
    EAF = effect_allele_frequency,
    
    # Odds ratio
    OR = odds_ratio,
    
    # 95% CI (derived from SE of log OR)
    OR_95U = exp(log(odds_ratio) + 1.96 * standard_error),
    OR_95L = exp(log(odds_ratio) - 1.96 * standard_error),
    
    # Strand (assumed)
    STRAND = "+",
    
    # Sample size (replace with real N if available)
    N = N ,
    
    # Chromosome
    CHROMOSOME = chromosome,
    
    # Position
    POSITION = base_pair_location
  ) %>%
  select(
    MARKERNAME,
    EA,
    NEA,
    EAF,
    OR,
    OR_95U,
    OR_95L,
    STRAND,
    N,
    CHROMOSOME,
    POSITION
  )

write.csv(meta1, "E:/Gwas_final/cat1_meta.csv", row.names = FALSE)
```


```{r}
#metastat_2

#metastat_2$N <- 300739  meta
#actual
meta2 <- metastat_2 %>%
  mutate(
    # Marker name
    MARKERNAME = hm_rsid,
    
    # Alleles (harmonised)
    EA = hm_effect_allele,
    NEA = hm_other_allele,
    
    # Effect allele frequency (if available separately)
    # Replace with correct column if you have hm_eaf
    EAF = effect_allele_frequency,
    
    # Odds ratio
    OR = exp(hm_beta),
    
    # Confidence intervals (already provided!)
    OR_95L = exp(hm_beta - 1.96 * standard_error),
    OR_95U = exp(hm_beta + 1.96 * standard_error),
    
    # Strand
    STRAND = "+",
    
    # Sample size (replace properly)
    N = N,
    
    # Chromosome & position (if available)
    CHROMOSOME = chromosome,
    POSITION = base_pair_location
  ) %>%
  select(
    MARKERNAME,
    EA,
    NEA,
    EAF,
    OR,
    OR_95U,
    OR_95L,
    STRAND,
    N,
    CHROMOSOME,
    POSITION
  )
write.csv(meta2, "E:/Gwas_final/cat2_meta.csv", row.names = FALSE)
```


```{r}
#adding sample size
metasta_3$N <- 61581
#main analy
meta3 <- metasta_3 %>%
  mutate(
    # Marker name
    MARKERNAME = hm_rsid,
    
    # Alleles (harmonised)
    EA = hm_effect_allele,
    NEA = hm_other_allele,
    
    # Effect allele frequency (if available separately)
    # Replace with correct column if you have hm_eaf
    EAF = effect_allele_frequency,
    
    # Odds ratio
    OR = exp(hm_beta),
    
    # Confidence intervals (already provided!)
    OR_95L = exp(hm_beta - 1.96 * standard_error),
    OR_95U = exp(hm_beta + 1.96 * standard_error),
    
    # Strand
    STRAND = "+",
    
    # Sample size (replace properly)
    N = N,
    
    # Chromosome & position (if available)
    CHROMOSOME = chromosome,
    POSITION = base_pair_location
  ) %>%
  select(
    MARKERNAME,
    EA,
    NEA,
    EAF,
    OR,
    OR_95U,
    OR_95L,
    STRAND,
    N,
    CHROMOSOME,
    POSITION
  )

write.csv(meta3, "E:/Gwas_final/cat3_meta.csv", row.names = FALSE)
```


```{r}
#add sample size
metasta_4$N <- 247540
#actual analysis
meta4 <- metasta_4 %>%
  mutate(
    # Marker name
    MARKERNAME = hm_rsid, 
    
    # Alleles (harmonised)
    EA = hm_effect_allele,
    NEA = hm_other_allele,
    
    # Effect allele frequency (if available separately)
    # Replace with correct column if you have hm_eaf
    EAF = hm_effect_allele_frequency,
    
    # Odds ratio
    OR = exp(hm_beta),
    
    # Confidence intervals (already provided!)
    OR_95L = exp(hm_beta - 1.96 * standard_error),
    OR_95U = exp(hm_beta + 1.96 * standard_error),
    
    # Strand
    STRAND = "+",
    
    # Sample size (replace properly)
    N = N,
    
    # Chromosome & position (if available)
    CHROMOSOME = hm_chrom,
    POSITION = hm_pos
  ) %>%
  select(
    MARKERNAME,
    EA,
    NEA,
    EAF,
    OR,
    OR_95U,
    OR_95L,
    STRAND,  
    N,
    CHROMOSOME,
    POSITION
  )
write.csv(meta4, "E:/Gwas_final/cat4_meta.csv", row.names = FALSE)

```



```{r}

#load gwama results following running on linux computer

gwama_out <- read_table("E:/Gwas_final/gwama_final/gwama.out")

```


```{r}
# ploting a manhattan plot from gwama results
# selecting chromosomes and positions

sele_gwas2_col <- Logpc23_100[, c("CHR", "SNP", "BP", "P", "OR", "beta")]


#rename the column in filtered_sig to SNP to match gwas column
colnames(gwama_out)[colnames(gwama_out) == "rs_number"] <- "SNP"

#merge the data based on SNP
merged_gwas2 <- merge(sele_gwas2_col, gwama_out, by = "SNP", all.x = TRUE)

#assigning true significant values
merged_gwas2$is_significant <- ifelse(merged_gwas2$`p-value` < 0.05, TRUE, FALSE)

# Update 'is_significant' to check if 'meta p value' is less than 0.05 and not NA
merged_gwas2$is_significant_meta <- ifelse(!is.na(merged_gwas2$`p-value`) & 
merged_gwas2$`p-value` < 0.05, TRUE, FALSE)

#select pontential significant SNPs after meta analysis(IMPORTANT)
significant_snps <- merged_gwas2[merged_gwas2$is_significant_meta == TRUE, ]

```

```{r}
#preparing to plot manhattan

merged_gwas2$`p-value`[is.na(merged_gwas2$`p-value`)] <- 0

#create a data frame for new plot

metaan <- merged_gwas2 %>% 
  
  # Compute chromosome size
  group_by(CHR) %>% 
  summarise(chr_len=max(BP)) %>% 
  
  # Calculate cumulative position of each chromosome
  mutate(tot=cumsum(chr_len)-chr_len) %>%
  select(-chr_len) %>%
  
  # Add this info to the initial dataset
  left_join(merged_gwas2, ., by=c("CHR"="CHR")) %>%
  
  # Add a cumulative position of each SNP
  arrange(CHR, BP) %>%
  mutate( BPcum=BP+tot) %>%

  # Add highlight and annotation information
  mutate(is_highlight = ifelse(-log10(`p-value`) > 7.3, "yes", "no")) %>%
  mutate( is_annotate=ifelse(-log10(`p-value`)>9, "yes", "no")) 

# Prepare X axis
axisdf <- metaan %>% group_by(CHR) %>% summarize(center=( max(BPcum) + min(BPcum) ) / 2 )

```


```{r}
ggplot(metaan2, aes(x = BPcum, y = -log10(`p-value`))) +
  geom_point(aes(color = as.factor(CHR)), alpha = 0.8, size = 1.3) +
  scale_color_manual(values = rep(c("brown", "pink"), 22)) +
  scale_x_continuous(labels = axisdf$CHR, breaks = axisdf$center) +
  scale_y_continuous(expand = c(0, 0),limits = c(1, 12.5)) +
  geom_hline(yintercept = 7.3, color = "red", linetype = "dashed") +  # e
  geom_hline(yintercept = upper_line, color = "blue", linetype = "dashed") + #lowerlimit
  theme_bw() +
   # Add highlighted points
  #geom_point(data=subset(metaan2, is_highlight=="yes"), color="orange", size=2) +
  
    # Add label using ggrepel to avoid overlapping
  geom_label_repel( data=subset(metaan2, is_annotate=="yes"), aes(label=SNP), size=2) +
  
  theme(
    legend.position = "none",
    panel.border = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  xlab("Genomic Position") +
  ylab("-log10(p-value)")


```


### working on heterogeneity

```{r}
Gwama_data <- gwama_out %>% filter(n_studies == 5)

Gwama_data$lower <- Gwama_data$beta - 1.96 * Gwama_data$se
Gwama_data$upper <- Gwama_data$beta + 1.96 * Gwama_data$se
```

```{r}
#selecting significant SNPs
Gwama_data$is_significant <- ifelse(Gwama_data$`p-value` < 0.05, TRUE, FALSE)
Meta_sign <- Gwama_data[Gwama_data$is_significant == TRUE, ]
pont_sig <- Meta_sign[Meta_sign$`p-value` < 5e-8, ]

```


```{r}
#Chcking for heterogeneity among significant SNPs
library(metafor)
res <- rma(yi = pont_sig$beta, sei = pont_sig$se, method = "REML")

summary(res)
```

```{R}

#selcting SNPs based on heterogeneity
homogen_snps <- pont_sig %>%
  filter(`q_p-value` >= 0.05 & i2 <= 0.50)

heterogen_snps <- pont_sig %>%
  filter(`q_p-value` < 0.05 & i2 > 0.50)


```

```{r}
#creating a forest plot
```{r}
forest(
  final_res2,
  transf = exp,
  refline = 1,
  slab = heterogen_snps$rs_number,
  xlab = "Effect Size (OR)",
  main = "Forest Plot for Heterogeneous SNPs"
)

mtext(
  paste0("Heterogeneity: Q = ", round(Q,2),
         ", p = ", format.pval(Qp, digits = 2),
         ", I² = ", round(I2,1), "%",
         ", τ² = ", round(tau2,3)),
  side = 1,
  line = 3,
  cex = 0.8
)

```

#following heterogeneity significance,  rerunning the analysis with a random effect or excluding the heterogeneous SNps can be done. we opted for a rerun using random effect model.



